---
title: "Reproducible Research: Course Project 2"
author: 'MD AHSANUL ISLAM'
date: "08/01/2021"
output:
  html_document:
    keep_md: yes
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cerulean
    # default, cerulean, journal, flatly, darkly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, and yeti
    highlight: haddock
    # default, tango, pygments, kate, monochrome, espresso, zenburn, haddock, breezedark, and textmate
    code_download: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "", prompt = TRUE, message=F, warning = F
)
```
---

# Synopsis


# Introduction


This is the second course project for the course "Reproducible Research" which is part of the Coursera’s "Data Science: Foundations using R" Specialization.

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data Processing

There is also some documentation of the database available:

- [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## Loading data

```{r, cache = T}
# bz2file <- bzfile("repdata_data_StormData.csv.bz2")
# if(!exists("storm")) {
#   storm <- read.csv(bz2file, na.strings = c("NA"))
# }

storm <- data.table::fread(input = "repdata_data_StormData.csv.bz2")
```


Loading required packages:
```{r}
library(tidyverse)
```


## Examining the data

```{r}
names(storm)
```


```{r}
dplyr::glimpse(storm)
```

```{r}
dim(storm)
```

## Variables to study

The variables that I will use to get some insights are as follows:

1. FATALITIES : approximate number of deaths   
2. INJURIES   : approximate number of injuries  
3. PROPDMG    : approximate property damags  
4. PROPDMGEXP : the units for property damage value  
5. CROPDMG    : approximate crop damages  
6. CROPDMGEXP : the units for crop damage value  
7. EVTYPE     : weather events

## Data Processing 

```{r}
df <- storm %>% select(FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
                       CROPDMG, CROPDMGEXP, EVTYPE) %>% 
  mutate_if(is.character, as.factor)
```

```{r}
glimpse(df)
```

Checking for missing values:
```{r}
summary(df)
```
There are missing values in `PROPDMGEXP` and `CROPDMGEXP` variables. For now we may ignore this. If any issue is faced then I may exclude these observations for the specific case.

### Health Impact - Data Processing

```{r, cache = T}
df.injuries <- df %>% 
  select(EVTYPE, INJURIES) %>% 
  group_by(EVTYPE) %>% 
  summarise(total.injuries = sum(INJURIES)) %>% 
  arrange(-total.injuries)
```

Top 10 events according to number of injuries:
```{r}
head(df.injuries, 10)
```

```{r, cache = T}
df.fatalities <- df %>% 
  select(EVTYPE, FATALITIES) %>% 
  group_by(EVTYPE) %>% 
  summarise(total.fatalities = sum(FATALITIES)) %>% 
  arrange(-total.fatalities)
```

Top 10 events according to number of fatalities/deaths:
```{r}
head(df.fatalities, 10)
```



### Economic Impact - Data Processing

According to [this](https://rpubs.com/gcctang1/271126/), the data provides two types of economic impact; Property damage (PROPDMG) and crop damage (CROPDMG). 

The actual damage in $USD is indicated by PROPDMGEXP and CROPDMGEXP parameters. The index in the PROPDMGEXP and CROPDMGEXP can be interpreted as the following:-

* H, h -> hundreds = x100  
* K, K -> kilos = x1,000    
* M, m -> millions = x1,000,000  
* B,b -> billions = x1,000,000,000  
* (+) -> x1  
* (-) -> x0  
* (?) -> x0  
* blank -> x0  
 
The total damage caused by each event type needs to be calculated manually - 

```{r}
df.damage <- df %>% select(EVTYPE,
                           PROPDMG, PROPDMGEXP,
                           CROPDMG, CROPDMGEXP) %>% 
  group_by(EVTYPE)
```

```{r, cache = T}
df.damage <- df %>% select(EVTYPE,
                           PROPDMG, PROPDMGEXP,
                           CROPDMG, CROPDMGEXP)
Symbol <- sort(unique(df.damage$PROPDMGEXP))
Multiplier <- c(0,0,0,1,10,10,10,10,10,10,10,10,10,10^9,10^2,10^2,10^3,10^6,10^6)
convert.Multiplier <- data.frame(Symbol, Multiplier)
df.damage.amount <- df.damage %>% 
  mutate(Prop.Multiplier = Multiplier[match(df.damage$PROPDMGEXP,
                                            convert.Multiplier$Symbol)],
         Crop.Multiplier = Multiplier[match(df.damage$CROPDMGEXP,
                                            convert.Multiplier$Symbol)],
         PROPDMG.amount = PROPDMG*Prop.Multiplier,
         CROPDMG.amount = CROPDMG*Crop.Multiplier,
         Total.damage = PROPDMG.amount + CROPDMG.amount) %>%
  group_by(EVTYPE) %>% 
  summarize(TOTAL.DMG.EVTYPE = sum(Total.damage))%>%
  arrange(-TOTAL.DMG.EVTYPE)
```

To 10 events according to amount of damage in USD:
```{r}
head(df.damage.amount, 10)
```

# Visulizations 

## Weather Events and Health

```{r}
df.injuries[1:10,] %>% ggplot(aes(x = total.injuries,
                                  y = reorder(EVTYPE, total.injuries))
                                ) +
  geom_bar(stat = "identity", fill = "#d00000") +
  scale_x_continuous(position = "top", breaks = seq(0, 90000, by = 10000),
                     expand = c(0,0)) +
  expand_limits(x = 100000) +
  geom_text(aes(label = total.injuries), colour = "black", hjust = 0) +
  labs(x = "Total Fatalities", y = "Weather Event Type",
       title = "Top 10 Events with Highest Total Injuries") +
  theme(plot.title = element_text(colour = "#370617", size = 18,
                                  face = "bold"),
        panel.background = element_blank(),
        panel.grid.major.x = element_line("grey"))
```



```{r}
df.fatalities[1:10,] %>% ggplot(aes(x = total.fatalities,
                                    y = reorder(EVTYPE, total.fatalities))
                                ) +
  geom_bar(stat = "identity", fill = "#6a040f") +
  scale_x_continuous(position = "top", breaks = seq(0, 6000, by = 800),
                     expand = c(0,0)) +
  expand_limits(x = 6000) +
  geom_text(aes(label = total.fatalities), colour = "black", hjust = 0) +
  labs(x = "Total Fatalities", y = "Weather Event Type",
       title = "Top 10 Events with Highest Total Fatalities") +
  theme(plot.title = element_text(colour = "#370617", size = 18,
                                  face = "bold"),
        panel.background = element_blank(),
        panel.grid.major.x = element_line("grey"))
```


## Weather events and economy 

```{r}

```










